Project Equalizer

### 2007-01-04 WhitePapers/ProjectEqualizer/paper.tex
website: Update project description and feature list

\maketitle
\thispagestyle{empty}
% \begin{figure}[ht]
% \centering
% \includegraphics[width=5em]{logo.pdf}
% \end{figure}
\vfill
\abstract{
% what
Equalizer is a project to develop software to simplify the creation of scalable
graphics applications and to improve the usability of multipipe
visualization systems. The main components of Equalizer are a resource
server, which controls the visualization system's configuration, a
client side library for the development of scalable graphics software, a
transparent software layer to execute unmodified applications alongside
with scalable applications, as well as remote visualization
capabilities. Equalizer is designed to support all kinds of
visualization systems ranging from laptops, workstations, multipipe
shared memory systems to graphics clusters.

@@

% how/when
The open development approach of Equalizer maximizes the benefits for
hardware vendors, software developers, research institutions and end
users. Interested parties are strongly encouraged to contact us. We are
looking for contributions or founding for several components of Equalizer.
}
\vfill
{\center\begin{tabularx}{.88\textwidth}{|l|l|X|}
\hline
\bf Version & \bf Date & \bf Changes \\
\hline
0.2         & October 5, 2006 & update status\\
0.1         & July 5, 2006    & initial draft\\
\hline
\multicolumn{3}{c}{\small Latest version at 
  \htmladdnormallink{http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}
  {http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}}\\
\end{tabularx}}
\clearpage
\pagenumbering{arabic}

\section{Overview}
As outlined in \cite{Analysis}, there are three fundamentally different
approaches to parallel rendering. Transparent solutions offer an easy
solution for applications which cannot or need not to be modified, but
are limited in performance and compatibility. Distributed
scene graphs offer better performance, but require the application
to use a certain scene graph. The third approach, a parallel rendering
framework offers a generic API for the development of parallel, scalable
graphics applications.

Today we have satisfying, though often proprietary, solutions for the
transparent approach\footnote{e.g., Chromium from Stanford University,
VGP from ModViz, SVN from IBM}. Distributed scene graphs are beginning
to emerge\footnote{e.g., ScaleViz from Mercury Computer Systems} as
well. There is currently no general programming framework to create
parallel graphics applications. Furthermore, all existing solutions are
'islands', that is, they are configured separately and do possibly
interfere with each other when running simultanously on the same
system. Equalizer addresses these issues.

@@

The Equalizer development is conducted in an open way. It is licensed
under the LGPL open source license. Contributing parties, such as
hardware vendors, software developers and end users, have the following
benefits from contributing to the project:

\begin{description}
\item[Lower Development Costs] The common functionality of any multipipe
  application is outsourced to the Equalizer framework, thus freeing the
  software developers to reimplement basic algorithms. The expertise for
  high performance visualization is partly embedded into Equalizer, and
  has not to be developed inhouse. The open development approach ensures
  that Equalizer delivers the functionality most needed by the community.
\item[Higher Functionality] Additional features, which would not be
  cost-effective to implement in the application, are provided by the
  Equalizer framework. New features will become available over time,
  and will often be usable by the application with no or minimal code
  changes.
\item[Ease of Use] The central resource management of Equalizer requires
  only a one-time setup, thus minimizing the impact for end users. Once
  a visualization system is set up, unmodified and scalable OpenGL
  applications can be run without any modification. Simple
  configurations, especially shared memory systems, will be configured
  automatically.
\item[Reduced Development Risk] The open source license of Equalizer is
  the guarantee for software developers that the product will be
  available indefinitely. Critical features or bugs, in case they are
  not addressed by the maintainers, can still be addressed internally.
\end{description}

@@

The programming framework is the main contribution of the
Equalizer project. It addresses common problems when creating parallel
graphics applications, and thus facilitates the development of scalable
multipipe rendering software. It addresses the process creation and
synchronization, rendering task distribution, data transport, and the
composition of the rendering results. The software developer plugs the
application's rendering code into the framework. The Equalizer server
deploys this rendering code according to the system's configuration and
load. The programming framework is the client library to the Equalizer
server, as well as the header files for this library.

@@

Much of the development of Equalizer is driven by the actual need of the
users, which are at the moment the software developers.

Currently, the programming framework is usable for the development of
commercial multipipe applications by an experienced OpenGL
programmer. By the end of 2006, we will have implemented the core
scalable rendering engine to show the benefits of scalable rendering on
commodity graphics clusters.

The transparent layer will be developed in 2007. The founding for this
effort is currently investigated. Interested parties are strongly
encouraged to contact us with proposals.

\subsection{Programming Framework}
The core part of the Equalizer project is the programming framework and
resource server. Naturally, these two components are the first major
milestone. Currently, they are usable to start the development of
parallel graphics applications. Right now Equalizer supports multipipe
rendering for display walls and immersive virtual reality installations,
including stereo and head tracking.

>>>>

\maketitle
\thispagestyle{empty}
% \begin{figure}[ht]
% \centering
% \includegraphics[width=5em]{logo.pdf}
% \end{figure}
\vfill
\abstract{
% what
Equalizer is a project to develop software to simplify the creation of scalable
graphics applications and to improve the usability of multipipe
visualization systems. The main components of Equalizer are a resource
server, which controls the visualization system's configuration, a
client-side library for the development of scalable graphics software, a
transparent software layer to execute unmodified applications alongside
with scalable applications, as well as remote visualization
capabilities. Equalizer is designed to support all kinds of
visualization systems ranging from laptops, workstations, multipipe
shared memory systems to graphics clusters.

@@

% how/when
The open development approach of Equalizer maximizes the benefits for
hardware vendors, software developers, research institutions and end
users. Interested parties are strongly encouraged to contact us. We are
looking for contributions or founding for several components of Equalizer.
}
\vfill
{\center\begin{tabularx}{.88\textwidth}{|l|l|X|}
\hline
\bf Version & \bf Date & \bf Changes \\
\hline
0.3         & January 4, 2007 & update status wrt 0.2 release, some rewording\\
0.2         & October 5, 2006 & update status\\
0.1         & July 5, 2006    & initial draft\\
\hline
\multicolumn{3}{c}{\small Latest version at 
  \htmladdnormallink{http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}
  {http://www.equalizergraphics.com/documents/ProjectEqualizer.pdf}}\\
\end{tabularx}}
\clearpage
\pagenumbering{arabic}

\section{Overview}
As outlined in \cite{Analysis}, there are three fundamentally different
approaches to parallel rendering. Transparent solutions offer an easy
solution for applications which cannot or need not to be modified, but
are limited in performance and compatibility. Distributed
scene graphs offer better performance, but require the application
to use a certain scene graph and are therefore relatively invasive. The
third approach, a parallel rendering framework offers a minimally
invasive approach for the development of parallel, scalable graphics
applications.

Today we have satisfying, though often proprietary, solutions for the
transparent approach\footnote{e.g., Chromium from Stanford University,
VGP from ModViz, SVN from IBM}. Distributed scene graphs are beginning
to emerge\footnote{e.g., OpenSG, ScaleViz from Mercury Computer Systems} as
well. There is currently no general programming framework to create
parallel graphics applications. Furthermore, all existing solutions are
'islands', that is, they are configured separately and do possibly
interfere with each other when running simultanously on the same
system. Equalizer addresses these issues.

@@

The Equalizer development is conducted in an open way. It is licensed
under the LGPL open source license, allowing the creation of commercial
and open source applications using the framework, while ensuring a
healthy community around the project. Contributing parties, such as
hardware vendors, software developers and end users, have the following
benefits from contributing to the project:

\begin{description}
\item[Lower Development Costs] The common functionality of any multipipe
  application is outsourced to the Equalizer framework, thus freeing the
  software developers to reimplement basic algorithms. The expertise for
  high performance visualization is partly embedded into Equalizer, and
  has not to be developed inhouse. The open development approach ensures
  that Equalizer delivers the functionality most needed by the community.
\item[Higher Functionality] Additional features, which would not be
  cost-effective to implement in the application, are provided by the
  Equalizer framework. New features will become available over time,
  and will often be usable by the application with no or minimal code
  changes.
\item[Ease of Use] The central resource management of Equalizer requires
  only a one-time setup, thus minimizing the impact for end users. Once
  a visualization system is set up, unmodified and scalable OpenGL
  applications can be run without any modification. Simple
  configurations, especially shared memory systems, will be configured
  automatically.
\item[Reduced Development Risk] The open source license of Equalizer is
  the guarantee for software developers that the product will be
  available 'forever'. Critical features or bugs, in the event that they are
  not addressed by the maintainers, can still be addressed internally.
\end{description}

@@

The programming framework is the main contribution of the
Equalizer project. It addresses common problems when creating parallel
graphics applications, and thus facilitates the development of scalable
multipipe rendering software. Equalizer provides functionality in the
area of process creation and synchronization, rendering task
distribution, data transport, as well as the composition of the
rendering results. The software developer plugs the application's
rendering code into the framework. The Equalizer server deploys this
rendering code according to the system's configuration and load. The
programming framework is the client library to the Equalizer server, as
well as the header files for this library.

@@

Much of the development of Equalizer is driven by the actual need of the
users, which are at the moment the software developers. A current
feature list can be found in \cite{Features}.

Currently, the programming framework is usable for the development of
commercial multipipe applications by an experienced OpenGL
programmer. Version 0.2 added support for scalable rendering to show the
benefits of scalable rendering on commodity graphics clusters and shared
memory systems.

For 2007, the main areas we plan to address are the extension of the
resource server for multi-application usage, general reliability and
performance improvements, a Windows port, as well as the integration of
a transparent layer. The founding for this effort is currently
investigated. Interested parties are strongly encouraged to contact us
with proposals.

\subsection{Programming Framework}
The core part of the Equalizer project is the programming framework and
resource server. Naturally, these two components are the first major
milestone. Currently, they are usable to start the development of
parallel graphics applications. Right now Equalizer supports multipipe
rendering for display walls and immersive virtual reality installations,
including stereo and head tracking, as well as a wide range of scalable
rendering features.
